---
title: "HARVARD EXTENSION SCHOOL"
subtitle: "CSCI E-106 - Data Modeling - Final Project"
author: 
 - Kaleo Mungin, Luciano Carvalho, Ibrahim Hashim, 
 - Bethun Bhowmik, Mohanish Kashiwar, Seymur Hasanov
tags: [House Price Prediction, Linear Regression, Neural Networks, Decision Trees, Support Vector Machines (SVM), Data Analysis, Model Selection, Statistical Testing, King County Real Estate, Predictive Analytics]
abstract: 
 - This project, undertaken by a team of students at Harvard Extension School, focuses on developing a comprehensive predictive model for house prices in King County, USA, using the 'KC_House_Sales' dataset. The dataset provides a rich variety of house attributes, allowing for an in-depth analysis of factors influencing property values. The initial phase of the project involves data cleaning and transformation, including the removal of irrelevant variables and conversion of data types. The primary analytical approach combines traditional linear regression models with more complex methods such as neural networks, decision trees, and support vector machines (SVM). The models are trained on a 70% split of the dataset and validated on the remaining 30%, ensuring robustness and accuracy in predictions.
 - In the subsequent stages, the project delves into graphical analysis, revealing key correlations and trends in the data. Techniques such as box plots, scatter plots, and heatmaps provide insights into the relationships between house prices and attributes like square footage, number of bedrooms, and location. The project also explores the impact of categorical variables and property age on pricing. Model performance is thoroughly tested using various metrics, including MSE and R-squared values. The final selection of the primary model, referred to as the "champion" model, is based on its performance on both the training and testing datasets. The project concludes with a discussion on model limitations, assumptions, and an ongoing monitoring plan, ensuring the model's relevance and accuracy over time.
date: "`r format(Sys.time(), '%d %B %Y')`"
geometry: margin=1.3cm
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---
\newpage
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Helper to display regression function with n coefficients
dispRegFunc <- function(reg) {
  coefs <- reg$coefficients
  b0 = coefs[1]
  n <- length(coefs)
  my_formula <- paste0("Y = ", round(b0, digits = 6))
  for (i in 2:n) {
    my_formula <- paste0(my_formula, " + ", round(coefs[i],6), names(coefs)[i])
  }
  my_formula
}

source("libs_funcs.R")

HouseSales<-read.csv("KC_House_Sales.csv")
```
\newpage
# Executive Summary

This section will describe the model usage, your conclusions and any regulatory and internal requirements. In a real world scenario, this section is for senior management who do not need to know the details. They need to know high level (the purpose of the model, limitations of the model and any issues).


\newpage
# I. Introduction (5 points)

*This section needs to introduce the reader to the problem to be resolved, the purpose, and the scope of the statistical testing applied. What you are doing with your prediction? What is the purpose of the model? What methods were trained on the data, how large is the test sample, and how did you build the model?*

--- in-progress ---

This project aims to develop a predictive model for house prices in King County, USA. The model's purpose is to provide an analytical tool for understanding the key factors influencing property values in this area, which is significant for various stakeholders in the real estate sector.

Our primary data source is the 'KC_House_Sales' dataset, which includes detailed information on various house attributes. **(Add here the hypothesis of initial exploration on house prices)**.

The methodology adopted for this project combines conventional statistical techniques with modern data analytics methods. Initial models will be based on linear regression, with further exploration into more complex approaches like neural networks and decision trees, depending on their performance in preliminary analyses.

To ensure robust model training and validation, the dataset has been divided into a training set (70%) and a test set (30%). This division is crucial for evaluating the model's effectiveness in making predictions on new, unseen data.

\newpage
# II. Description of the data and quality (15 points)

*Here you need to review your data, the statistical test applied to understand the predictors and the response and how are they correlated. Extensive graph analysis is recommended. Is the data continuous, or categorical, do any transformation needed? Do you need dummies? *

## Data Overview

```{r Data Overview}

df_house = read.csv("KC_House_Sales.csv")
cat("Number of NA values in dataframe:",sum(is.na(df_house))) #no NA values
head(df_house)

```



## Data Types, Categories and Cleaning

[Add here info about the data and what was performed]

```{r Data types, Categories and Cleaning}
# [Kaleo]
#removing `id` column
df_house = subset(df_house, select = -id)

#converting `price` to numeric
df_house$price <- as.numeric(gsub("[\\$,]", "", df_house$price))

##cleaning `date` and adding `year`,`month`,`day` columns
df_house$date <- as.POSIXct(df_house$date, format = "%Y%m%d")
df_house$year <- as.numeric(format(df_house$date, "%Y"))
df_house$month <- as.numeric(format(df_house$date, "%m"))
df_house$day <- as.numeric(format(df_house$date, "%d"))

head(df_house)
```

```{r Convert variables to numeric data}
# [Kaleo]
#converting all to numeric...ignores date column
ndf_house = df_house[sapply(df_house, is.numeric)]
# use df_house instead, unless you have a specific reason!
head(ndf_house)

```

## Stat Summary - in progress

[Add here the initial analysis of the summary, if applicable]

```{r Stat Summary}

str(ndf_house)
summary(ndf_house)

```

```{r Stat summary of the price}
summary(ndf_house$price)
```


## Graphical Analysis - in progress - Seymur

The pairwise scatter plot shows the correlation between highly correlated variables extracted from heatmap correlation matrix.

- there is a strong correlation between sqft_living and price, indicated that as the living area increases, so does the house price increase.
- A strong positive correlation with grade, implying that higher-quality houses tend to be more expensive.
- Each variable's distribution is shown on the diagonal, with price notably skewed towards lower values, indicating most homes are on the more affordable end of the spectrum with fewer high-priced outliers.

```{r Pairwise scatter plot}

# Pairwise scatter plot with correlation coefficients
ggpairs(ndf_house, columns = c("price", "sqft_living", "bedrooms", "bathrooms", "grade"), 
        title = "Pairwise Scatter Plots with House Price")

```

**Analysis**: The histogram below shows the distribution of house prices, revealing that most houses are in the lower price range with a significant decrease in the number of houses as the price increases, indicating a right-skewed distribution with relatively few high-priced houses.

```{r Histogram of prices}

# Histogram of house prices
ggplot(df_house, aes(x = price)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(title = "Histogram of House Prices", x = "Price", y = "Count")

```
**Analysis**: The first boxplot shows that house prices generally increase with the number of bedrooms, but vary widely, especially for homes with more bedrooms. The second boxplot indicates that house prices rise with better house grades, with greater price variability at higher grades.Both plots demonstrate that while there is a general trend of increasing price with more bedrooms and higher grades, there is also a considerable variation within these categories. The presence of outliers suggests that factors other than the number of bedrooms and house grade can significantly influence house prices.

```{r Outlier analysis using Boxplots}

# Boxplot for price by number of bedrooms
ggplot(df_house, aes(x = factor(bedrooms), y = price)) + 
  geom_boxplot() +
  labs(title = "Boxplot of Prices by Number of Bedrooms", x = "Number of Bedrooms", y = "Price")

# Boxplot for price by house grade
ggplot(df_house, aes(x = factor(grade), y = price)) + 
  geom_boxplot() +
  labs(title = "Boxplot of Prices by House Grade", x = "Grade", y = "Price")

```
**Analysis**: the average price tends to increase with the number of bedrooms up to a certain point, but then the trend is less consistent. For instance, homes with 6 bedrooms have a higher average price than those with 7 or 8 bedrooms, and the average price for homes with 11 bedrooms is lower than for those with fewer bedrooms. There is a notable outlier at 33 bedrooms with a relatively low average price, which could indicate an atypical property or data error.

```{r Average Price analysis}

average_prices <- aggregate(price ~ bedrooms, data = ndf_house, FUN = mean)
ggplot(average_prices, aes(x = factor(bedrooms), y = price)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Average Price by Number of Bedrooms",
       x = "Number of Bedrooms",
       y = "Average Price") +
  theme_minimal()

```

```{r }

ggplot(df_house, aes(x = sqft_living, y = price)) +
  geom_point(aes(color = factor(bedrooms)), alpha = 0.6) +
  scale_color_brewer(type = 'seq', palette = 'Blues') +
  labs(title = "House Price vs. Square Footage of Living Space", x = "Square Footage of Living Space", y = "Price")

```

```{r Outlier analysis for Waterfront vs Mountain view}

ggplot(df_house, aes(x = factor(waterfront), y = price)) +
  geom_boxplot() +
  labs(title = "Boxplot of House Prices by Waterfront", x = "Waterfront", y = "Price")

```

**Analysis**: From geographical map, it is evident that properties situated around the latitude line of 47.6 and longitude between -122.25 and -122.00, which corresponds to the central and northern parts of Seattle, command higher prices per square foot. The relatively lower-priced properties per square foot, shown in orange, are more dispersed and located primarily south of central Seattle, extending towards Tacoma, as well as in the outlying suburban areas.It is also noticeable that along the latitudinal line around 47.4, there are pockets of high-priced properties per square foot, potentially indicating affluent neighborhoods or areas with high-value real estate.

The map clearly shows a correlation between location and property value per square foot, with central urban areas exhibiting the highest values. This pattern is typical for urban centers where proximity to amenities, employment opportunities, and other socioeconomic factors drive up real estate prices.

```{r Geo dist of house prices in King County}

# Scatter plot of properties
ggplot(data = ndf_house, aes(x = long, y = lat, color = price)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Geographical Distribution of House Prices in King County",
       x = "Longitude", y = "Latitude", color = "Price") +
  theme_minimal()

```

## [Analysis section] 

From geographical map, it is evident that properties situated around the latitude line of 47.6 and longitude between -122.25 and -122.00, which corresponds to the central and northern parts of Seattle, command higher prices per square foot. The relatively lower-priced properties per square foot, shown in orange, are more dispersed and located primarily south of central Seattle, extending towards Tacoma, as well as in the outlying suburban areas.It is also noticeable that along the latitudinal line around 47.4, there are pockets of high-priced properties per square foot, potentially indicating affluent neighborhoods or areas with high-value real estate.

The map clearly shows a correlation between location and property value per square foot, with central urban areas exhibiting the highest values. This pattern is typical for urban centers where proximity to amenities, employment opportunities, and other socioeconomic factors drive up real estate prices.

```{r Map visualizatoin of average prices by squre foot distribution over long and lat}

register_stadiamaps(key = '765cbcd1-2ec2-40e3-8a81-98624e616208')
df_summary <- ndf_house %>%
  mutate(price_per_sqft = price / sqft_living) %>%
  group_by(long, lat) %>%
  summarize(avg_price_per_sqft = mean(price_per_sqft, na.rm = TRUE), .groups = 'drop')

king_county_map <- get_stadiamap(bbox = c(left = min(df_house$long), 
                                         bottom = min(df_house$lat), 
                                         right = max(df_house$long), 
                                         top = max(df_house$lat)),
                                 maptype = "stamen_terrain",
                                 zoom = 10)
ggmap(king_county_map) +
  geom_point(data = df_summary, aes(x = long, y = lat, color = avg_price_per_sqft), 
             alpha = 0.5, size = 1) +
  scale_color_gradient(low = "orange", high = "darkblue", name = "Avg Price per Sqft") +
  labs(title = "Average Price per Sqft of Properties in King County",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```
## Zipcode visualizations (can discard)
***Zipcode is a valid predictor***

```{r}
# [Kaleo]
ggplot( df_house, aes(x = as.factor(zipcode), y = price)) +
  geom_boxplot()+theme(axis.text.x = element_text(angle = 90,vjust=0.5))
```

```{r}
# [Kaleo]
# Prices by Zipcode heatmap
heatmap_data <- df_house %>%
  group_by(zipcode) %>%
  summarize(med_price = median(price),
            min_price = min(price),
            max_price = max(price))
heatmap_data <- heatmap_data %>% 
  arrange(desc(med_price))

ggplot(heatmap_data, aes(x = factor(zipcode, levels = zipcode), y = 1, fill = med_price)) +
  geom_tile() +
  geom_text(aes(label = paste(scales::comma(min_price),"/",
                              scales::comma(med_price),"/",
                              scales::comma(max_price)),
                vjust = 0.5,angle=90),
            color = "black", size = 3)+
  scale_fill_gradient(low = "lightyellow", high = "red", name = "Median Price",
                      labels = scales::comma_format()) +
  labs(title = "Heatmap of median prices by zipcode (min/med/max)",
       x = "zip",
       y = "") +
  theme(axis.text.x = element_text(angle = 90, hjust = 0,vjust=0.5),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "bottom",
        legend.key.width = unit(2, "cm"))
```

## Correlation Analysis - in progress

***[Add here the initial analysis of the correlation matrix]***

```{r Correlation Matrix}
# [Kaleo]
#correlation matrix
cor_matrix = cor(ndf_house)
correlation_df = as.data.frame(cor_matrix)

#new dataframe with variables 'highly' (>0.2) correlated with price
x_high = subset(ndf_house, select = c(view,waterfront,sqft_basement,sqft_living,
                                      sqft_living15,sqft_above,grade,bathrooms,bedrooms,floors,lat))

pheatmap(cor_matrix,
         color = colorRampPalette(c("blue", "white", "red"))(20),
         main = "correlation matrix heatmap",
         fontsize = 8,
         cellwidth = 15,
         cellheight = 11,
         display_numbers = TRUE
)
```

**Creating a new dataframe with variables 'highly' (>0.2) correlated with price**

```{r}
# [Kaleo]
head(x_high)
```

## Data Transformation

## Categorical Variables and Age Analysis

### Renovation Indicator Variable

A binary variable named 'renovated' was introduced to indicate whether a property has undergone renovation. This variable is set to 1 if the 'yr_renovated' field is not zero, signifying that the property has been renovated at least once. Otherwise, it is set to 0, indicating no renovation. This distinction provides a straightforward way to assess the impact of renovations on property values.

```{r}

# Creating a new variable 'renovated' 
# 1 if the property has been renovated (yr_renovated != 0), 0 otherwise
df_house$renovated = ifelse(df_house$yr_renovated != 0, 1, 0)

```

### Property Age Calculation

This is a tentative way to consider "age since las renovation" and see if there's a correlation between the time a house was last built/renovated on its selling price. A new variable called 'age' was calculated to represent the current age of each property. If a property was renovated, its age is the difference between 2023 and the renovation year ('yr_renovated'). If not renovated, the age is the difference between 2023 and the year the house was built ('yr_built'). This variable helps in understanding the effect of property age and recent renovations on house prices.

```{r}

# Creating 'age' column
df_house$age = ifelse(df_house$renovated == 1, 2023 - df_house$yr_renovated, 2023 - df_house$yr_built)

head(df_house)

```


```{r}
df_house = df_house[-c(1, 13)]

set.seed(1023)
n<-dim(df_house)[1]
IND<-sample(c(1:n),round(n*0.7))
train.dat<-df_house[IND,]
test.dat<-df_house[-c(IND),]

dim(train.dat)
dim(test.dat)
```


## Summary of the section II

\newpage
# III. Model Development Process (15 points) - (Bethun Bhowmik)

*Build a regression model to predict price.  And of course,  create the train data set which contains 70% of the data and use set.seed (1023). The remaining 30% will be your test data set. Investigate the data and combine the level of categorical variables if needed and drop variables. For example, you can drop id, Latitude, Longitude, etc. *

The KC House sales dataset is split into train(70%) and test(30%) datasets, with 15129 observations in the train dataset and 6484 observations in the test dataset.

```{r}
#[Bethun]
set.seed(1023)
n<-dim(df_house)[1]
IND<-sample(c(1:n),round(n*0.7))
train.dat<-df_house[IND,]
test.dat<-df_house[-c(IND),]

dim(train.dat)
dim(test.dat)
```

**Analysis**:
The house_lm model to predict price has:
- Residual Standard Error of 193,600
- Multiple R-square of 71.38%
- p-value of < 2.2e-16
- MSE of 37,494,980,807

Overall, the model appears to be statistically significant overall given the low p-value for the F-statistic. The multiple R-square suggests that 71.38% of the variability in price is explained by the model.
Moreover, all most predictors (except sqft_lot) have a statistically significant impact on the house price. Predictions on the test dataset should be done to further validate model usefulness.

From the model plot we observe the following:
- LINEARITY ASSUMPTION: From the Residuals vs Fitted plot, as the residuals seem to be randomly dispersed around the horizontal axis and the line is horizontal, a linear regression model may be appropriate, however we may need to transforming the response variable.
- NORMALITY ASSUMPTION: From the Q-Q Residuals plot, as the points are out of the diagonal line on either ends, the residuals do not meet the normality assumption.
- CONSTANT VARIANCE ASSUMPTION: The scale location plot points to non constant variance. The HOMOSCEDASTICITY ASSUMPTION is violated.Moreover the line is curvilinear.
- The impact of bad data like influential and outlier observations seems to be impactful when inspecting the the Residual vs. Leverage graph

```{r}
#[Bethun]
house_lm<-lm(price ~.,data=train.dat)
summary(house_lm)
par(mfrow=c(2,2))
plot(house_lm)
MSE <- summary(house_lm)$sigma^2
print(MSE)
```

**Analysis**:
The R-square values on both training and test data are quite close - 71.38% and 70.86% respectively. This suggests that the model generalizes well to new data, maintaining a similar level of explanatory power. The RMSE of 204,868.3 indicates the average magnitude of errors in predicting house prices on the test data. The SSE of 2.721402e+14 represents the sum of squared differences between predicted and observed values on the test data. A lower RMSE and SSE indicates better model fit, and we will explore furher on how these values can be lowered.

```{r}
#[Bethun]
# Test data Predictions
house_lm_test_pred <- predict(house_lm, newdata = test.dat)

house_lm_test_mse <- mean((house_lm_test_pred - test.dat$price)^2)
house_lm_test_rmse <- sqrt(house_lm_test_mse)
house_lm_test_residuals <- test.dat$price - house_lm_test_pred
house_lm_test_rsq <- 1 - var(house_lm_test_residuals) / var(test.dat$price)
house_lm_test_sse <- sum((test.dat$price - house_lm_test_pred)^2)
results.df <- data.frame(model = "Linear Regression Test Data Predictions",
                         R.Squared.Train = summary(house_lm)$r.square,
                         R.Squared.Test = house_lm_test_rsq,
                         RMSE.test = house_lm_test_rmse,
                         SSE.test = house_lm_test_sse)
print(results.df)
```

**Analysis**:
The insignificant predictor sqft_lot is dropped and the model is refitted. We observe that the Multiple R-square value remains unchanged at 71.38%. The model is as follows:

Price = -74344803.92767 + -34633.405255bedrooms + 32821.338262bathrooms + 143.671881sqft_living + 8767.015515floors + 531351.549043waterfront + 52065.87946view + 24875.912771condition + 82737.693562grade + 22.301049sqft_above + -896.223982yr_built + 4065.130653yr_renovated + -507.797776zipcode + 473638.095898lat + -206204.537141long + 21.65297sqft_living15 + -0.25171sqft_lot15 + 38482.930296year + 1380.790546month + -331.741298day + -7982223.272107renovated + 1440.530732age + 101303.883852price_binary

```{r}
#[Bethun]
predictors_to_drop <- c("sqft_lot")

# Update the model by excluding the specified predictors
updated <- as.formula(paste("price ~ .", paste0("- `", predictors_to_drop, "`", collapse = "")))
house_lm <- update(house_lm, formula = updated)
summary(house_lm)
dispRegFunc(house_lm)
```


**Analysis**:

On plotting the boxplot of residuals, we observe multiple outliers on both ends of the whiskers. Moreover, when we plot the residuals against the fitted values, we observe that the residuals spread out with increase in fitted values along x-axis.

```{r}
#[Bethun]
ei<-house_lm$residuals
boxplot(ei,horizontal=TRUE,staplewex=0.5,col=2,xlab="House Price Regression Residuals")
plot(house_lm$fitted.values,ei,xlab="Fitted Values House Linear Model",ylab="Residuals House Linear Model")
```

**Analysis**:
Predictors with low p-values of < 2.2e-16 (e.g. bedrooms, bathrooms, sqft_living, etc.) are statistically significant and are important in predicting price. Some predictors that have a have higher p-value (e.g. floors, zipcode, month, day) may not be statistically significant in predicting price.

```{r}
#[Bethun]
anova(house_lm)
```



**Analysis**:

A two sides t statistical test is used for testing individual coefficients and their significance. The critical t -value applicable given alpha=0.01 and n-2 degrees of freedom is 2.576.
H0: intercept =0, HA: intercept is not 0
H0: slope= 0    , HA: slope is not 0
Decision rule: when t > t critical reject null

The model, as indicated by the significant coefficients, suggests that features such as the number of bedrooms, bathrooms, square footage of living space, waterfront status, view, grade, etc, play a statistically significant role in predicting the house price.

```{r}
#[Bethun]

conf= 0.01/2     #Note we divide by 2 because this is a two tail test,if alpha=0.05 then use .05/2
df<-21613-2            #manually calculating the degrees of freedom
value<-formatC(qt(conf,df,lower.tail=FALSE))    
print(paste("Critical T values: ",value))
#coefficients
matrix_coef <- summary(house_lm)$coefficients  # Extract coefficients in matrix
matrix_coef   

my_estimates <- matrix_coef[ , 1]  # Matrix manipulation to extract estimates

#Step 6: Pr(>|t|) < 0.01 there is sufficient statistical evidence to reject null for both parameters. Using a t-test we noted that t-values (6.259865 and abs(4.102897) are larger than t-critical that is 2.637
```


$H_o$: Error variances are constant
$H_a$: Error variances are not constant
Decision Rule is if statistic> critical reject the null
or if p-value < alpha (0.01) reject the hull

P value is < 2.2e-16. So we reject $H_o$, Error variances are not constant.

**Analysis**:

```{r}
#[Bethun]
bptest(house_lm, studentize = FALSE)
```
**Analysis**:
For Boxcox transormation, the optimal lambda is -0.04040404 with 95% confidence interval range which is close to zero.Although, we can estimate lambda to be zero and apply logarithmic transformation of the data, we have done the boxcox transformation with exact optimal lambda value for better accuracy.

```{r}
#[Bethun]
par(mfrow=c(1,1))
bc<-boxcox(house_lm,lambda=seq(-4,4,by=0.1))
lambda <- bc$x[which.max(bc$y)] 
lambda  #This is the optimal lambda
```
**Analysis**:
After boxcox transformation, we observe that
- Residual Standard Error has reduced significantly to 0.004981 from 193,600
- Multiple R-square has increased to 84.03% from 71.38%
- p-value remains same at < 2.2e-16

Overall, the model appears to be a much better fit that before. From the model plot we observe the following:
- LINEARITY ASSUMPTION: The Residuals vs Fitted plot looks better than before and confirms that a linear regression model is appropriate
- NORMALITY ASSUMPTION: The points are much better aligned along the diagonal in the Q-Q Residuals plot, however some tails remain
- CONSTANT VARIANCE ASSUMPTION: The Scale Location plot points to constant variance
- There are still some outlier observations which may be to be impactful as seen from the the Residual vs. Leverage graph

```{r}
#[Bethun]
house_lm1<-lm(price^lambda~.,data=train.dat)
summary(house_lm1)
par(mfrow=c(2,2))
plot(house_lm1)
dispRegFunc(house_lm1)
```
**Analysis**:
On predicting the values of the Boxcox transformed model on the test dataset, we observe a R-square test of 75.79% compared to R-square train of 84.03%, from which we conclude that the transformed model generalizes well to new data.

```{r}
#[Bethun]
pred<- predict(house_lm1,test.dat)^(1/lambda)
act<-test.dat$price

SST<-var(act)*(length(act)-1)
SSE<-sum((act-pred)^2)
R.Square.Test <- 1- SSE/SST
R.Square.Test
R.Square.Train=0.8403

knitr::kable(cbind(R.Square.Train,R.Square.Test), align = "c",
      caption = "Pseudo $R^2$ values for Train and Test")
```

\newpage
# III. Model Development Process (15 points) - SVM Mohanish (in progress)

*Build a regression model to predict price.  And of course,  create the train data set which contains 70% of the data and use set.seed (1023). The remaining 30% will be your test data set. Investigate the data and combine the level of categorical variables if needed and drop variables. For example, you can drop id, Latitude, Longitude, etc. *



```{r}

# find optimal cost of misclassification
# tune.out <- tune(svm, price~., data = train.dat, kernel = "linear",
#                 ranges = list(cost = c(0.001, 0.01, 0.1, 1, 5, 10, 100)))
# extract the best model
# (bestmod <- tune.out$best.model)
```


```{r}

```

```{r}

```


```{r}

```

```{r}

```


```{r}

```


```{r}

```


\newpage
# IV. Model Performance Testing (15 points)

*Use the test data set to assess the model performances. Here, build the best multiple linear models by using the stepwise both ways selection method. Compare the performance of the best two linear models. Make sure that model assumption(s) are checked for the final linear model. Apply remedy measures (transformation, etc.) that helps satisfy the assumptions. In particular you must deeply investigate unequal variances and multicollinearity. If necessary, apply remedial methods (WLS, Ridge, Elastic Net, Lasso, etc.). *



## Best Subset Regression

```{r}

house_lm2<-lm(price^lambda~.,data=test.dat)
# k1<-ols_step_best_subset(house_lm2)
# k1
# plot(k1,guide="none")
```





\newpage
# V. Challenger Models (15 points)

*Build an alternative model based on one of the following approaches to predict price: regression tree, NN, or SVM. Explore using a logistic regression. Check the applicable model assumptions. Apply in-sample and out-of-sample testing, backtesting and review the comparative goodness of fit of the candidate models. Describe step by step your procedure to get to the best model and why you believe it is fit for purpose.*


## Random Forest

```{r}
#[Kaleo]

ntree_values <- c(55, 112, 150, 200, 250)

mse_values = numeric(length(ntree_values))
test_rsq_values = numeric(length(ntree_values))
train_rsq_values = numeric(length(ntree_values))

# commented out cross validation to save computation time!!

# for (i in seq_along(ntree_values)) {
#   ntree = ntree_values[i]
#   
#   model = randomForest(price ~ ., data = train.dat, ntree = ntree)
#   predictions_test <- predict(model, newdata = test.dat)
#   predictions_train <- predict(model, newdata = train.dat)
#   
#   mse_values[i] <- mean((predictions_test - test.dat$price)^2)
#   test_rsq_values[i] = cor(predictions_test,test.dat$price)^2
#   train_rsq_values[i] = cor(predictions_train,train.dat$price)^2
# }
```

```{r}
#[Kaleo]
mse_hard.coded = c(18959180557,19044061189,19147650224,18975362291,18893426330)
test_rsq_values_hard.coded = c(0.8734077,0.8730189,0.8719060,0.8740622,0.8747472)

# Plot the results
plot(ntree_values, mse_hard.coded, type = "b", xlab = "Number of Trees", ylab = "Mean Squared Error", main = "Random Forest Cross Validation Test MSE (ntrees)")

plot(ntree_values, test_rsq_values_hard.coded,type="o", xlab = "Number of Trees", ylab = "Test R-squared", main = "Random Forest Cross Validation Test R-squared (ntrees)")
text(ntree_values,test_rsq_values_hard.coded,labels=round(test_rsq_values_hard.coded,4),pos=3,xpd=TRUE)
```
```{r}
# [Kaleo]

#commented out for computation time..

#using efficient model (50 trees...for now)
house_rf = randomForest(price ~ ., data = train.dat, ntree = 50)
rf_test_pred = predict(house_rf,newdata=test.dat)
rf_train_pred = predict(house_rf,newdata=train.dat)

rf_test_mse = mean((rf_test_pred - test.dat$price)^2)
rf_test_rsq = cor(rf_test_pred,test.dat$price)^2
rf_train_rsq = cor(rf_train_pred,train.dat$price)^2

cat("test mse:",rf_test_mse,"train rsq:", rf_train_rsq, "test rsq:",rf_test_rsq)
```

```{r}
#[Kaleo]

# relative variable importance
par(mar = c(7, 5, 4, 2))

var_importance = importance(house_rf)
top_features = var_importance[order(var_importance[,1], decreasing = TRUE), ][1:12]

barplot(top_features,names.arg = rownames(top_features),las=2)
```


## Regression Tree Model

```{r}
# [Luciano]

# Build the regression tree model
tree_model <- rpart(price ~ ., data = train.dat, method = "anova")
print(summary(tree_model))

rpart.plot(tree_model, main="Regression Tree", extra=1, under=TRUE, faclen=0, cex=0.75)

# Prediction and Performance
tree_predictions <- predict(tree_model, test.dat)
tree_rmse <- sqrt(mean((tree_predictions - test.dat$price)^2))
print(paste("Tree RMSE:", tree_rmse))

```


## Neural Network Model

```{r}
# [Luciano]

# Build the neural network model
nn_model <- nnet(price ~ ., data = train.dat, size = 5, linout = TRUE, trace = FALSE, MaxNWts = 5000)
print(nn_model)

# Prediction and Performance
nn_predictions <- predict(nn_model, test.dat, type = "raw")
nn_rmse <- sqrt(mean((nn_predictions - test.dat$price)^2))
print(paste("NN RMSE:", nn_rmse))

```


## Support Vector Machine (SVM) Model

```{r}
# [Luciano]

# Build the SVM model
svm_model <- svm(price ~ ., data = train.dat)
print(summary(svm_model))

# Prediction and Performance
svm_predictions <- predict(svm_model, test.dat)
svm_rmse <- sqrt(mean((svm_predictions - test.dat$price)^2))
print(paste("SVM RMSE:", svm_rmse))

```


## Logistic Regression Model

```{r Price binary calculation }
# [Luciano]

# Calculate the median price
median_price <- median(train.dat$price)

# Create a binary variable based on the median price
train.dat$price_binary <- ifelse(train.dat$price > median_price, 1, 0)
test.dat$price_binary <- ifelse(test.dat$price > median_price, 1, 0)

```

```{r Logistic Regression Model and Confusion Matrix}
# [Luciano]

log_model <- glm(price_binary ~ ., data = train.dat, family = binomial())
print(summary(log_model))

# Predicting probabilities
log_predictions_prob <- predict(log_model, test.dat, type = "response")

# Binning probabilities into two categories: 0 and 1
log_predictions <- ifelse(log_predictions_prob > 0.5, 1, 0)

# Ensuring factors have the same levels
log_predictions_factor <- factor(log_predictions, levels = c(0, 1))
test_price_binary_factor <- factor(test.dat$price_binary, levels = c(0, 1))

# Confusion Matrix
log_conf_mat <- caret::confusionMatrix(log_predictions_factor, test_price_binary_factor)
print(log_conf_mat)

```

``` {r Model Diagnostics for Logistic Model}

# Diagnostic Plots for the Logistic Regression
par(mfrow = c(2, 2))
plot(log_model)

```
4 UPs

**Analysis**: The diagnostic plots suggest that while the regression model captures the overall trend, there are indications of non-normality in residuals, potential heteroscedasticity, and the presence of influential outliers, warranting further investigation into data transformation or alternative modeling approaches.


```{r ROC Curve for Logistic Regression}

# Assuming binary classification
roc_curve <- pROC::roc(test.dat$price_binary, log_predictions)
plot(roc_curve, main = "ROC Curve")

```
ROC Curve


**Analysis**: The ROC curve demonstrates perfect classification performance with maximal sensitivity and specificity


```{r Variable Importance for Tree Model}

# randomForest model for variable importance
rf_model <- randomForest(price ~ ., data = train.dat)
randomForest::varImpPlot(rf_model)

```
Random Forest plot

**Analysis**: The Variable Importance plot from a Random Forest model highlights 'price.binary' as the most influential predictor, with 'sqft_living', 'grade', and 'sqft_living15' also contributing significantly to model accuracy, suggesting their strong predictive power in the Random Forest model.



\newpage
# VI. Model Limitation and Assumptions (15 points)

*Based on the performances on both train and test data sets, determine your primary (champion) model and the other model which would be your benchmark model. Validate your models using the test sample. Do the residuals look normal? Does it matter given your technique? How is the prediction performance using Pseudo R^2, SSE, RMSE?  Benchmark the model against alternatives. How good is the relative fit? Are there any serious violations of the model assumptions? Has the model had issues or limitations that the user must know? (Which assumptions are needed to support the Champion model?)* 

```{r}
# [Kaleo]
# Results dataframe...We're supposed to use Pseudo R-squared, SSE, RMSE, as seen above. 
# We'll have to look into 'Pseudo R-squared' most likely

results.df
```
\newpage
# VII. Ongoing Model Monitoring Plan (5 points)

*How would you picture the model needing to be monitored, which quantitative thresholds and triggers would you set to decide when the model needs to be replaced? What are the assumptions that the model must comply with for its continuous use?*


\newpage
# VIII. Conclusion (5 points)

*Summarize your results here. What is the best model for the data and why?*

# Bibliography (7 points)

*Please include all references, articles and papers in this section.*

# Appendix (3 points)

*Please add any additional supporting graphs, plots and data analysis.*


